---
title: "MSc Bioinformatics"
author: "Marc Henrion"
date: "17 January 2020"
output: pdf_document
---


```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=16, fig.height=9, dpi=150, highlight=T)

require(tidyverse)
require(knitr)
require(gridExtra)
```


# Session 7 - Practical (Solutions)


1. Using the `iris` dataset (type `?iris` to get more information about this dataset) that comes pre-loaded with R, produce the following figures:
    + Produce histograms for each of `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width`. Try using both the graphics (base R) and ggplot2 (tidyverse R) libraries.
    + Produce a bar plot for `Species`.
    + Produce box and whisker plots for each of the 4 continuous variables. Put them all on a single, multi-panel figure.
    + Repeat for just `Sepal.Length` using a violin plot, stratifying by `Species`.
    + Produce a single graph (not multi-panel) that has histograms for `Sepal.Length` for each of the 3 flower species.
    + There are 4 continuous variables. This means there are 6 possible pairs of these. For each such pair, produce a scatter plot of one variable against the other and highlight the different flower species by using a different colour for each species.
    + For one of these 6 scatter plots: estimate the bivariate probability density and add density contour lines to the figure.
    
* Histograms using base R:

```{r}
par(mfrow=c(2,2))
for(i in 1:4){
  hist(iris[,i],
       breaks=25,
       main="Histogram",
       xlab=gsub(pattern="\\.",replacement=" ",colnames(iris)[i]))
}
```

* Histograms using ggplot2:

```{r}
g<-list()
g[[1]]<-ggplot(data=iris,mapping=aes(x=Sepal.Length)) + geom_histogram(bins=25)
g[[2]]<-ggplot(data=iris,mapping=aes(x=Sepal.Width)) + geom_histogram(bins=25)
g[[3]]<-ggplot(data=iris,mapping=aes(x=Petal.Length)) + geom_histogram(bins=25)
g[[4]]<-ggplot(data=iris,mapping=aes(x=Petal.Width)) + geom_histogram(bins=25)
grid.arrange(g[[1]],g[[2]],g[[3]],g[[4]],nrow=2)
```

* Barplot

```{r}
barplot(table(iris$Species),
        main="distribution of flower species in the iris dataset",
        ylab="frequency")
```

* Box plots

```{r}
par(mfrow=c(1,4))
for(i in 1:4) {
  boxplot(iris[,i], main=names(iris)[i])
}
```

* Violin plot of `Sepal.Length`

```{r}
ggplot(data=iris,mapping=aes(x=Species,y=Sepal.Length,fill=Species)) +
  geom_violin() +
  geom_boxplot(width=0.05, fill="white")
```

* Histograms for `Sepal.Length`

```{r}
ggplot(data=iris,mapping=aes(x=Sepal.Length,fill=Species)) +
  geom_histogram(binwidth=0.15,position="dodge")
```

* Pair-wise scatterplots

```{r}
par(mfrow=c(2,3))
for(i in 1:3){
  for(j in min(c(i+1),4):4){
    print(c(i,j))
    plot(iris[,i],
         iris[,j],
         pch=20,
         col=ifelse(iris$Species=="setosa","salmon",ifelse(iris$Species=="versicolor","darkgreen","steelblue")),
         xlab=colnames(iris)[i],
         ylab=colnames(iris)[j])
  }
}
```

* Bivariate density contours

```{r}
library(MASS)
clrs<-colorRampPalette(c("blue","red","orange","yellow","white"))
z <- kde2d(iris$Sepal.Length, iris$Petal.Width, n=50)
plot(iris[,c("Sepal.Length","Petal.Width")],
     xlab="Sepal.Length",
     ylab="Petal.Width",
     col=ifelse(iris$Species=="setosa","salmon",ifelse(iris$Species=="versicolor","darkgreen","steelblue")),
     pch=20)
contour(z, drawlabels=FALSE, nlevels=11,col=clrs(14), add=TRUE)
```


2. Install the package `nycflights13`, then load it. This has data on flights that took off in the US during 2013. There are 5 data tables:
    + `airlines`, data on airlines
    + `airports`, data on airports
    + `planes`, data on planes
    + `weather`, hourly weather data at NYC airports for 2013
    + `flights`, data on flights leaving NYC airports during 2013

  * Compute the average delay by destination, then join the airports data frame to get the longitude and latitude of delays. Plot this (if you are using ggplot2, then the functions `borders()` and `coord_quickmap()` can be useful for a nicer figure).
    
  * Construct data frames giving average delay per wind speed / temperature / precipitation / visibility. Produce scatter plots of each of these against delay and add an average trend line.

```{r, warning=F, message=F}
library(nycflights13)

# Compute the average delay by destination, then join the airports data frame
# to get the longitude and latitude of delays.
avg_dest_delays <-
	flights %>%
	group_by(dest) %>% 
	summarise(delay = mean(arr_delay, na.rm = TRUE)) %>% # arrival delay NA's are cancelled flights
	inner_join(airports, by = c(dest = "faa"))

# stratify by origin airport
avg_dest_delays_by_origin <-
	flights %>%
	group_by(origin,dest) %>% 
	summarise(delay = mean(arr_delay, na.rm = TRUE)) %>% # arrival delay NA's are cancelled flights
	inner_join(airports, by = c(dest = "faa"))

# plotting this
ggplot(data=avg_dest_delays,mapping=aes(lon,lat,colour=delay,size=delay+20)) +
  borders("state") +
  geom_point() +
  coord_quickmap(xlim=c(-130,-70),ylim=c(20,55)) # xlim, ylim to hide Alaska and Hawaii
```

```{r, warning=F, message=F}
flights_weather <-flights %>% left_join(weather,by=c("year","month","day","hour"))

flights_precip <- flights_weather %>% 
  group_by(precip) %>%
  summarise(delay=mean(dep_delay,na.rm=T))

flights_wind <- flights_weather %>%
  group_by(wind_speed) %>%
  summarise(delay=mean(dep_delay,na.rm=T))

flights_temp <- flights_weather %>%
  group_by(temp) %>%
  summarise(delay=mean(dep_delay,na.rm=T))

flights_visib <- flights_weather %>%
  group_by(visib) %>%
  summarise(delay=mean(dep_delay,na.rm=T))

g1<-ggplot(data=flights_precip,mapping=aes(x=precip,y=delay)) +
  geom_point() +
  geom_smooth()

g2<-ggplot(data=flights_wind,mapping=aes(x=wind_speed,y=delay)) +
  geom_point() +
  geom_smooth()

g3<-ggplot(data=flights_temp,mapping=aes(x=temp,y=delay)) +
  geom_point() +
  geom_smooth()

g4<-ggplot(data=flights_visib,mapping=aes(x=visib,y=delay)) +
  geom_point() +
  geom_smooth()

grid.arrange(g1,g2,g3,g4)
```
